Web-based Supplementary Materials for Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Steve De Rosa, Mario Roederer and Raphael Gottardo

### 2012-07-27 MIMOSA.tex
Continued previous changes
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations. Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV-1-stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial~. We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000, and 50,000 events, with ten independent realizations of data for each $N$. The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change. This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D). In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} A,B).

For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Web Figure B, panel A). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Web Figure B panel B).
%Since the constrained model relies on Monte-Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

@@

When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression. Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B). Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. A solution would be to explore alternative model parametrizations that could be used to reduce the number of required parameters. For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.

>>>>

We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations. Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV-1-stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial. We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000 events, with ten independent realizations of data for each $N$. The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change. This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C). In addition, the nominal \textit{vs.} observed FDR was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).

For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity evan at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F). Finally, even when MIMOSA was fit to data simulated from a model that violated model assumptions, it outperformed competing methods (Web Figure B, panels A-D).
%Since the constrained model relies on Monte-Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

@@

When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). 
In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression. 
Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B). 
Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes. 
In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. 
A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters. 
For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. 
As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. 
More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.

### 2012-07-27 WebSupplementaryMaterialMIMOSA.tex
Continued previous changes
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
%    \begin{scope} [x={(a.south east)},y={(a.north west)}]
%        \node at (0,1) [font=\tiny\sffamily] {A} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of A] (B) {
%    \begin{tikzpicture}
%    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
%        \begin{scope} [x={(b.south east)},y={(b.north west)}]
%        \node at (0,1) [font=\tiny\sffamily] {B} ;
%        \end{scope}
%    \end{tikzpicture}
%    };
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \node [below=of A] (C) {
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
    \begin{scope} [x={(e.south east)},y={(e.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {C} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of C] (D) {
    \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
        \begin{scope} [x={(f.south east)},y={(f.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {D} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [below=of C] (E) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
         \begin{scope} [x={(g.south east)},y={(g.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {E} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of E] (F) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
        \begin{scope} [x={(h.south east)},y={(h.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {F} ;
\end{scope}
    \end{tikzpicture}
    };
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two--sided data with small counts. Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.  A) The average ROC from 10 simulation with N=5,000 events. B) The average observed and nominal FDR from 10 simulations with N=5,000 events. C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events. MIMOSA fit to two--sided data were simulated from the standard model with N=1,000 events. A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations.}
   \label{webfig:simulations_trunc}
\end{figure}
\bibliographystyle{mn2e}
\bibliography{MIMOSA}

>>>>

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
%    \begin{scope} [x={(a.south east)},y={(a.north west)}]
%        \node at (0,1) [font=\tiny\sffamily] {A} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of A] (B) {
%    \begin{tikzpicture}
%    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
%        \begin{scope} [x={(b.south east)},y={(b.north west)}]
%        \node at (0,1) [font=\tiny\sffamily] {B} ;
%        \end{scope}
%    \end{tikzpicture}
%    };
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \node [below=of A] (C) {
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
    \begin{scope} [x={(e.south east)},y={(e.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {C} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of C] (D) {
    \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
        \begin{scope} [x={(f.south east)},y={(f.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {D} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [below=of C] (E) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
         \begin{scope} [x={(g.south east)},y={(g.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {E} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [right=of E] (F) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
        \begin{scope} [x={(h.south east)},y={(h.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {F} ;
\end{scope}
    \end{tikzpicture}
    };
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two--sided data with small counts. Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.  A) The average ROC from 10 simulation with N=5,000 events. B) The average observed and nominal FDR from 10 simulations with N=5,000 events. C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events. MIMOSA fit to two--sided data were simulated from the standard model with N=1,000 events. E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.}
   \label{webfig:simulations_trunc}
\end{figure}
\bibliographystyle{mn2e}
\bibliography{MIMOSA}
